---
title: "HFD - Proteo"
output:
  html_document: default
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE,warnings=F,error = F,include=T)

library(ggplot2)
library(ggforce)
library(limma)
library(dplyr)
library(ggrepel)
library(msigdbr)
library(imputeLCMD)
library(ggsci)
library(org.Mm.eg.db)
library(clusterProfiler)
library(factoextra)
library(emmeans)
library(patchwork)
library(pheatmap)
library(brglm2)
library(readxl)

#Convenience function for plotting the volcano plot and saving the results
plot_table_save = function(df){
  df$Compound = rownames(df)
  ix = match(df$Compound,data.wide$PG.ProteinNames)
  df$PG.Genes = data.wide$PG.Genes[ix]
  df$PG.ProteinDescriptions = data.wide$PG.ProteinDescriptions[ix]
  df$Significant = ifelse(df$adj.P.Val<0.05,"Significant","NS")
  df$label = ifelse(df$Significant == "NS","",df$Compound)
  df$Direction = ifelse(df$logFC>0,"Up regulated","Down regulated")
  df$Direction[df$Significant=="NS"] = "NS"
  
  volc = ggplot(df,aes(x=logFC,y=-log10(adj.P.Val),color=Direction)) + geom_point()+
  theme_bw(16) +facet_wrap(~Contrast)+geom_text_repel(data = df[df$label!= "",] ,aes(label=Compound))+scale_color_manual(values = c("Up regulated"="red","Down regulated" = "darkgreen","NS = Grey")) + ggtitle(df$Contrast)

  print(volc)
  
  ggsave(plot = volc,path=paste0(gsub("\\..*","", file)),filename = paste0(df$Contrast[1],"_volcano.png"),create.dir = T)
  print("\n")

knitr::kable(df[df$adj.P.Val<1,],row.names = F,digits = 4)

  write.csv(df,file =paste0(gsub("\\..*","", file),"/",df$Contrast[1],"_results.csv") )  
  
}


#Plots top 20 boxplots
plot_differences = function(df){
  plot.df = df
plot.df$Diet = factor(plot.df$Diet,levels=c("Con.diet","HFD"))
plot.df$Geno = factor(plot.df$Geno,levels=c("WT/WT","KI/KI"))
  print(ggplot( plot.df,aes(x = Geno,y=value,color=Diet)) +geom_boxplot(position = position_dodge()) + facet_wrap(~variable,scales="free_y")+theme_bw(16)+scale_color_aaas()+ggtitle(title))

  ggsave(path=paste0(gsub("\\..*","", file)),filename = paste0("","_top_boxplot.png"),create.dir = T)
  
}
 

#Plots top 20 boxplots
run_overrep = function(df){
  ix = match(df$Compound,data.wide$PG.ProteinNames)
  df$PG.Genes = data.wide$PG.Genes[ix]
  sig.proteins = unique(df[df$adj.P.Val<0.05,]$PG.Genes)
  if(length(sig.proteins) > 0 ){
    universe.proteins = unique(df$PG.Genes)
    universe.entrez = unique(na.omit(mapIds(org.Mm.eg.db, keys = universe.proteins,
           column = "ENTREZID", keytype = "SYMBOL")))
    sig.entrez = unique(na.omit(mapIds(org.Mm.eg.db, keys = sig.proteins,
           column = "ENTREZID", keytype = "SYMBOL")))
 
    enrich.wt = clusterProfiler::enrichGO(gene = sig.entrez, OrgDb = 'org.Mm.eg.db',keyType = "ENTREZID",universe=universe.entrez)
    p1 = barplot(enrich.wt, showCategory=20) 
    p2 = dotplot(enrich.wt, showCategory=20) 
    enrich.wt.net <- DOSE::setReadable(enrich.wt, 'org.Mm.eg.db', 'ENTREZID')
    p3 = cnetplot(enrich.wt.net, node_label="all") 
    print(cowplot::plot_grid( p2 , p3,ncol=2,rel_widths=c(0.7,1)))
    ggsave(path=paste0(gsub("\\..*","", file)),filename = paste0(df$Contrast[1],"_overrep_plots.png"),create.dir = T)

    path=paste0(gsub("\\..*","", file),"/",df$Contrast[1],"_overrep_results.csv")
    tmp.or = enrich.wt@result
    write.csv(tmp.or,file=path)
  } else {
    return("Not enough DE Proteins")
  } 
}
 
 
#Plots top 20 boxplots
run_gsea = function(df){
  ix = match(df$Compound,data.wide$PG.ProteinNames)
  df$PG.Genes = data.wide$PG.Genes[ix]
  proteins = df$PG.Genes
 
  logFC <- df$logFC
 
  names(logFC) <- proteins
  logFC = logFC[!duplicated(names(logFC))]
  logFC = logFC[!is.na(logFC)]
  logFC <- sort(logFC, decreasing = T)
  logFC = logFC[!is.na(names(logFC))]
  gsea = gseGO(logFC,ont ="BP", 
        keyType = "SYMBOL", 
        OrgDb =  'org.Mm.eg.db',
        pvalueCutoff = 0.2)

 
  if(nrow(gsea@result) > 0){
 
    p1 = enrichplot::dotplot(gsea, showCategory=20, split=".sign") + facet_grid(.~.sign)
    enrich.net <- DOSE::setReadable(gsea, 'org.Mm.eg.db', 'SYMBOL')
    p2 = cnetplot(enrich.net, node_label="all",foldChange=logFC) 
    # p3 = ridgeplot(gsea)  
    print(cowplot::plot_grid(p1, p2 ,ncol=3,rel_widths=c(0.7,1)))
    ggsave(path=paste0(gsub("\\..*","", file)),filename = paste0(df$Contrast[1],"_gsea_plots.png"),create.dir = T)
    path=paste0(gsub("\\..*","", file),"/",df$Contrast[1],"_gsea_results.csv")
    tmp.gsea = gsea@result
    write.csv(tmp.gsea,file=path)
  } else {
 
    return("No Enrichments with FDR < 0.2")
  }
}


```


# Section {.tabset}

```{r, include=F}
file = "~PC00064_YueMa_MouseLung_CompleteDataSet_Report_20240618_sent.xlsx"

file= "~/bbc-secondary/research/MOOD_yue_MS_template_VBCS1110/PC00064_YueMa_MouseLung_CompleteDataSet_Report_20240618_sent.xlsx"
## Please select the path to the data file you want to analyze
data = read_excel(file)

names = make.names(paste0(data$PG.ProteinNames),unique=T)
data.wide = data
data.wide$PG.ProteinNames = names
data = as.data.frame(t(subset(data[,12:27])))
colnames(data) =names
data$Sample = rownames(data)


# This line makes all of the compound variables numeric, sometimes they read in as characters
data[,1:(ncol(data)-1)] <- sapply( data[,1:(ncol(data)-1)], as.numeric )

Molecule = colnames(data)[1:(ncol(data)-1)]

data$Geno_Diet =  sapply(strsplit(data$Sample,"_"), function(x) x[3])
data$Geno = gsub("-.*","",data$Geno_Diet)
data$Geno = factor(data$Geno,levels=c("wt","ki"),labels = c("WT/WT","KI/KI"))
data$Diet = toupper(gsub(".*-","",gsub("-B.*","",data$Geno_Diet)))
data$Diet = factor(data$Diet,levels=c("LFD","HFD"),labels=c("Con.diet","HFD"))


table(data$Geno,data$Diet)

data = subset(data, select= - c(Geno_Diet))


```

## Missingness {.tabset}

Here we create a heatmap of missingness to see if it specific to any group of interest (Diet, Genotype). If there is any specificity, you would see very clear clustering blocks. If nothing shows up, there were no missing values


```{r, fig.height = 9,fig.width=15}
#Replace 0s with NA
data[data==0]=NA

miss.df = data
rownames(miss.df) = miss.df$Sample
miss.df = miss.df[,colnames(miss.df) %in% Molecule]
miss.df[!is.na(miss.df)] = 1
miss.df[is.na(miss.df)] = 0

missing = apply(miss.df,2, function(x) sum(x)/length(x))
anno = data.frame(Genotype = data$Geno,Diet = data$Diet)
rownames(anno) = rownames(miss.df)

if( length(which(missing<0.95)) > 0){
  pheatmap(t(miss.df[,which(missing<0.95)
  ]),color=c("white","black"),annotation_col = anno,show_rownames = F)
}

```


### Testing for specificity


```{r, include=F}

miss.df$Geno = data$Geno
miss.df$Diet = data$Diet
colnames(miss.df) = make.names(colnames(miss.df))
res.diet=NULL
res.geno=NULL
res.delta.delta=NULL
# 
# for(i in colnames(miss.df[,which(missing<0.95)])){
#   
#   f=as.formula(paste0(i,"~ Diet*Geno"))
#   fit = glm(f, data = miss.df, family = "binomial", method = brglmFit)
#   
#   res.diet = rbind(res.diet,data.frame(Compound = i,summary(emmeans(fit,pairwise~Diet|Geno,adjust="none")$contrasts)))
#   
#   res.geno = rbind(res.geno,data.frame(Compound = i,summary(emmeans(fit,pairwise~Geno|Diet,adjust="none")$contrasts)))
#   
#   res.delta.delta = rbind(res.delta.delta,data.frame(Compound = i,joint_tests(fit)[3,]))
#   res.delta.delta$model.term = "Delta - Delta"
# 
# }
# 
# 
# res.geno$FDR = p.adjust(res.geno$p.value,method="BH")
# 
# res.diet$FDR = p.adjust(res.diet$p.value,method="BH")
# 
# res.delta.delta$FDR = p.adjust(res.delta.delta$p.value,method="BH")
# 


```


```{r}

# if(length(colnames(miss.df[,which(missing<0.95)]))>0){
# 
#   knitr::kable(res.diet[res.diet$FDR < 0.1,],caption="Compounds specific to a Diet")
#   knitr::kable(res.geno[res.geno$FDR < 0.1,],caption="Compounds specific to a Genotype")
#   knitr::kable(res.delta.delta[res.delta.delta$FDR < 0.1,],caption="Compounds specific to a Diet/Genotype Combo")
# }


```

### Removed for too much missingness

```{r}

#These compounds are being removed for excessive missingness
knitr::kable(names(missing[missing<0.7]))

#drop heavily missing data (>30%)
data = data[, ! colnames(data) %in% names(missing[missing<0.7])]

```


```{r}

#impute
data[,colnames(data) %in% Molecule] = t(imputeLCMD::impute.QRILC(log2(t(data[,colnames(data) %in% Molecule])))[[1]])


```

## Exploratory {.tabset}
### PCA

```{r, fig.height=6,fig.width=16}

pc.plot = function(df){

  pc = prcomp(data.matrix(df[,colnames(df) %in% Molecule]))

  pc.df = as.data.frame(pc$x)
  pc.df$Genotype = df$Geno
  pc.df$Diet = df$Diet

  p1 = ggplot(pc.df,aes(x=PC1,y=PC2,color = Diet,shape=Genotype, linetype=Genotype )) +
    geom_point() +
    geom_mark_ellipse() +
    theme_bw(12) +
    scale_color_nejm()

  p2 = ggplot(pc.df,aes(x=PC3,y=PC4,color = Diet,shape=Genotype, linetype=Genotype )) +
    geom_point() +
    geom_mark_ellipse() +
    theme_bw(12) +
    scale_color_nejm()

  p3 = ggplot(pc.df,aes(x=PC5,y=PC6,color = Diet,shape=Genotype, linetype=Genotype )) +
    geom_point() +
    geom_mark_ellipse() +
    theme_bw(12) +
    scale_color_nejm()

  print(p1+p2+p3)

}

pc.d = pc.plot(data) 

ggsave(pc.d,file=paste0(gsub("\\..*","", file),"/pc.tif"))


```



### Heatmap {.tabset}

#### Pooled

```{r, fig.width=9,fig.height=25}

anno = data.frame(Genotype = data$Geno, data$Diet)
rownames(anno) = data$Sample
rownames(data) = data$Sample
pheatmap(t((data[,colnames(data) %in% Molecule])),
         annotation_col = anno,
         scale="row",
         color=colorRampPalette(c("blue","white","red"))(100),show_rownames = F)


```


## Differential Abundance {.tabset}

### Pooled {.tabset}

```{r, fig.height=14,fig.width=14}
#We have to drop the slashes to make syntactically correct names
data$Geno_Diet =  gsub(".*/","",paste0(data$Geno,"_",data$Diet))

design = model.matrix(~ 0 + Geno_Diet , data= data)


fit <- lmFit(t(data[,colnames(data) %in% Molecule]), design, robust = TRUE)


cont2 = makeContrasts( Geno_DietKI_HFD - Geno_DietWT_HFD ,
                       Geno_DietKI_Con.diet- Geno_DietWT_Con.diet,
                       Geno_DietWT_HFD - Geno_DietWT_Con.diet,
                       Geno_DietKI_HFD - Geno_DietKI_Con.diet,
                       levels = design)

set.seed(777)
fit.cont <- contrasts.fit(fit, contrasts = cont2) 

set.seed(777)
ebayesf <- eBayes(fit.cont, robust = T, trend = T)

tops.sig = topTable(ebayesf, adjust.method = "BH", p.value = 0.05, number = Inf)



tops.all = topTable(ebayesf, adjust.method = "BH", p.value = 1, number = Inf)

```

#### HFD KI/KI - WT/WT

```{r, fig.height=9,fig.width=9}

hfd.df = data.frame(Contrast = "KI/KI HFD vs WT/WT HFD", topTable(ebayesf, adjust.method = "BH", p.value = 1, number = Inf,coef=1))
hfd.df$Compound = rownames(hfd.df)
hfd.df$Significant = ifelse(hfd.df$adj.P.Val<0.05,"Significant","NS")
hfd.df$label = ifelse(hfd.df$Significant == "NS","",hfd.df$Compound)

# ggplot(hfd.df,aes(x=logFC,y=-log10(P.Value),color=Significant)) + geom_point()+
#   theme_bw(16) +facet_wrap(~Contrast)+geom_text_repel(data = hfd.df[hfd.df$label!= "",] ,aes(label=Compound))+scale_color_manual(values = c("NS"="black","Significant" = "red")) + ggtitle("HFD KI - WT")
# 
# knitr::kable(hfd.df[hfd.df$adj.P.Val<0.05,],row.names = F,digits = 4)

plot_table_save(hfd.df)
```

##### Over representation

```{r, fig.height=9,fig.width=25}

run_overrep(hfd.df)

```

##### GSEA

```{r, fig.height=9,fig.width=25}

run_gsea(hfd.df)

```

#### Con.diet KI/KI - WT/WT

```{r, fig.height=9,fig.width=9}

Con.diet.df = data.frame(Contrast = "KI/KI Con.diet vs WT/WT Con.diet", topTable(ebayesf, adjust.method = "BH", p.value = 1, number = Inf,coef=2))
Con.diet.df$Compound = rownames(Con.diet.df)
Con.diet.df$Significant = ifelse(Con.diet.df$adj.P.Val<0.05,"Significant","NS")
Con.diet.df$label = ifelse(Con.diet.df$Significant == "NS","",Con.diet.df$Compound)

# ggplot(Con.diet.df,aes(x=logFC,y=-log10(P.Value),color=Significant)) + geom_point()+
#   theme_bw(16) +facet_wrap(~Contrast)+geom_text_repel(data = Con.diet.df[Con.diet.df$label!= "",] ,aes(label=Compound))+scale_color_manual(values = c("NS"="black","Significant" = "red")) + ggtitle("Con.diet KI - WT")
# 
# knitr::kable(Con.diet.df[Con.diet.df$adj.P.Val<0.05,],row.names = F,digits = 4)

plot_table_save(Con.diet.df)
```


##### Over representation

```{r, fig.height=9,fig.width=25}

run_overrep(Con.diet.df)

```

##### GSEA

```{r, fig.height=9,fig.width=25}

run_gsea(Con.diet.df)

```

#### WT/WT HFD - Con.diet

```{r, fig.height=9,fig.width=9}

wt.df = data.frame(Contrast = "WT/WT hfd vs WT/WT/WT Con.diet", topTable(ebayesf, adjust.method = "BH", p.value = 1, number = Inf,coef=3))
wt.df$Compound = rownames(wt.df)
wt.df$Significant = ifelse(wt.df$adj.P.Val<0.05,"Significant","NS")
wt.df$label = ifelse(wt.df$Significant == "NS","",wt.df$Compound)

plot_table_save(wt.df)
```


##### Over representation

```{r, fig.height=9,fig.width=25}

run_overrep(wt.df)

```

##### GSEA

```{r, fig.height=9,fig.width=25}

run_gsea(wt.df)

```


#### KI/KI HFD - Con.diet

```{r, fig.height=9,fig.width=9}

ki.df = data.frame(Contrast = "KI/KI hfd vs KI/KI Con.diet", topTable(ebayesf, adjust.method = "BH", p.value = 1, number = Inf,coef=4))
ki.df$Compound = rownames(ki.df)
ki.df$Significant = ifelse(ki.df$adj.P.Val<0.05,"Significant","NS")
ki.df$label = ifelse(ki.df$Significant == "NS","",ki.df$Compound)
# 
# ggplot(Con.diet.df,aes(x=logFC,y=-log10(P.Value),color=Significant)) + geom_point()+
#   theme_bw(16) +facet_wrap(~Contrast)+geom_text_repel(data = Con.diet.df[Con.diet.df$label!= "",] ,aes(label=Compound))+scale_color_manual(values = c("NS"="black","Significant" = "red")) + ggtitle("KI HFD - Con.diet")
# 
# knitr::kable(Con.diet.df[Con.diet.df$adj.P.Val<0.05,],row.names = F,digits = 4)

plot_table_save(ki.df)

```


##### Over representation

```{r, fig.height=9,fig.width=25}

run_overrep(ki.df)

```

##### GSEA

```{r, fig.height=9,fig.width=25}

run_gsea(ki.df)

```

#### Boxplots Top Compounds

```{r, fig.height=12,fig.width=16}


  plot.df = reshape2::melt(data[,colnames(data) %in% c("Geno","Diet",rownames(tops.all)[1:20])],id.vars=c("Geno","Diet"))
  title = "Boxplots of Compounds with largest differences"

plot_differences(plot.df) 


```

