---
title: "Enrichment analysis"
author: "Daisy Fu"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    self_contained: yes
    toc: true
    toc_depth: 5
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: true
---

# Background

JIRA ticket number: VBCS-1202. This is rerun for only 4 comparisons. 

GO-BP, KEGG, four custom pathways, and reactome will be analyzed. 


Steps in the analysis:  

1. Import DE gene results

2. Run GSEA on GO_BP and KEGG

3. Plotting: dotplots for all.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=F, message=F, cache=F)
```

# Load packages


```{r load_libs, echo=TRUE, warning=F, message=F, cache=FALSE}

if (!"/varidata/research/projects/bbc/tools/Fu_R_4.4_ondemond_lib" %in% .libPaths()) {
  current_pathes <- .libPaths()
  .libPaths(c("/varidata/research/projects/bbc/tools/Fu_R_4.4_ondemond_lib", current_pathes))
}

orgdb <- "org.Mm.eg.db" 

set.seed(4321)

library(orgdb, character.only=TRUE)
library(SummarizedExperiment)
library(enrichplot)
library(tidyverse)
library(edgeR)
library(clusterProfiler)
library(msigdb)
library(cowplot)
library(readxl)
library(ReactomePA)

```

# Import DE results


```{r import_data}

# dir <- ("../rnaseq_workflow/results/")

dir.create("data_save_RDS")
dir.create("gsea_data")
dir.create("gsea_res")

pvalue = 1

# import custom gene sets:
list.files("../more_enrich_Mar_2025/from_client/")
canonical_senesce <- read_tsv("../more_enrich_Mar_2025/from_client/0_1 Canonical Senesce Pathway.txt") |>
  pull(pathway)
senescence_initiators <- read_tsv("../more_enrich_Mar_2025/from_client/0_2 Senescence Initiators.txt") |>
  pull(pathway)
senescence_response <- read_tsv("../more_enrich_Mar_2025/from_client/0_3 Senescence Responses.txt") |>
  pull(pathway)
senMayo_gene_mouse <- read_xlsx("../more_enrich_Mar_2025/from_client/SenMayo gene list 41467_2022_32552_MOESM4_ESM.xlsx", sheet = "mouse") |>
  pull( "Gene(murine)" )

gs_list <- list(canonical_senesce= canonical_senesce, 
                senescence_initiators = senescence_initiators, 
                senescence_response = senescence_response, 
               senMayo_gene_mouse = senMayo_gene_mouse )

custom_gs_dataset <- lapply(seq_along(gs_list), function(index) {
  df <- data.frame(gs_name = rep(names(gs_list)[index], length(gs_list[[index]])),
                   entrez_gene = unlist(gs_list[[index]]))
})

custom_gs_dataset_df <-  do.call(rbind.data.frame, custom_gs_dataset)


### ---- import DE gene results

read_DE_res_run_enrichment <- function(res_df = res_df, save_name = save_name) {

  gene_list <- sign(res_df$logFC) * (-log10(res_df$PValue))

  names(gene_list) <- res_df$entrez ## the names of the gene list is the entreze ID.

  if(any(is.na(gene_list))){
    gene_list = gene_list[-which(is.na(gene_list))]
  }
  if(any(is.na(names(gene_list)))) {
    gene_list = gene_list[-which(is.na(names(gene_list)))]
  }

  gene_list <- gene_list[!duplicated(names(gene_list))]

  gene_list <- sort(gene_list, decreasing = TRUE)


##### ----- run GO_BP first
## make sure the sorted/ranked comparison were in a list with names of comparisons.

  res_GOBP <-  gseGO(gene_list,
    OrgDb = org.Mm.eg.db,
    ont          = "BP",
    minGSSize    = 50,
    maxGSSize    = 500,
    pvalueCutoff = 0.05,
    verbose      = FALSE)
  
  y <- setReadable(res_GOBP, OrgDb = org.Mm.eg.db, keyType="ENTREZID")
  
  saveRDS(y, file = paste0("gsea_data/", save_name, "_gseGO_BP.rds"))
  write_tsv(y %>% data.frame(), file = paste0("gsea_res/", save_name, "_gseGO_BP.tsv"))
  
 
   ### next, run KEGG. 
  kk <- gseKEGG(geneList     = gene_list,
               organism     = 'mmu',
               minGSSize    = 50,
               pvalueCutoff = 0.05,
               verbose      = FALSE)

  z <- setReadable(kk, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
  
  saveRDS(z, file = paste0("gsea_data/", save_name, "_KEGG.rds"))
  write_tsv(z %>% data.frame(), file = paste0("gsea_res/", save_name, "_KEGG.tsv"))
  
  ## run reactome
  reactome_obj <- gsePathway(gene_list, 
                pvalueCutoff = 0.05,
                pAdjustMethod = "BH", 
                organism = "mouse",
                verbose = FALSE)
  
  reactome_obj <- setReadable(reactome_obj, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
  saveRDS(reactome_obj, file = paste0("gsea_data/", save_name, "_reactome.rds"))
  write_tsv(reactome_obj %>% data.frame(), file = paste0("gsea_res/", save_name, "_reactome.tsv"))
  
  
  # lastly, run gsea on the custom gene sets. 
  
  # we need to recreate a gene list with symbols
  gene_list_symbol <- sign(res_df$logFC) * (-log10(res_df$PValue))

  names(gene_list_symbol) <- res_df$Symbol ## the names of the gene list is the entreze ID.

  if(any(is.na(gene_list_symbol))){
    gene_list_symbol = gene_list_symbol[-which(is.na(gene_list_symbol))]
  }
  if(any(is.na(names(gene_list_symbol)))) {
    gene_list_symbol = gene_list_symbol[-which(is.na(names(gene_list_symbol)))]
  }

  gene_list_symbol <- gene_list_symbol[!duplicated(names(gene_list_symbol))]

  gene_list_symbol <- sort(gene_list_symbol, decreasing = TRUE)
  
  gsea_custom <- GSEA(gene_list_symbol, TERM2GENE = custom_gs_dataset_df, pvalueCutoff = pvalue, eps = 0, minGSSize = 15, nPermSimple = 10000)
  
  write_rds(gsea_custom, file = paste0("gsea_data/", save_name, "_custom_gene_sets", ".rds"))
  gsea_export <- gsea_custom %>% as.data.frame()
  write_tsv(gsea_export, file = paste0("gsea_res/", save_name, "_custom_gene_sets.tsv"))


  ## then do some plotting
  print(dotplot(gsea_custom, showCategory=20, split = ".sign", font.size = 10) + facet_grid(~.sign) +
            scale_y_discrete(labels=function(x) str_wrap(x, width=60)) +
        ggtitle(paste0("dotplot for gene set enrichment of four custom gene sets ", save_name)))
        
          
  ## then do some plotting
  print(dotplot(res_GOBP, showCategory=20, split = ".sign", font.size = 9) + facet_grid(~.sign) +
            scale_y_discrete(labels=function(x) str_wrap(x, width=60)) +
        ggtitle(paste0("dotplot for gene set enrichment of GO_BP - ", save_name)))
        
        
  print(dotplot(kk, showCategory=20, split = ".sign", font.size = 9) + facet_grid(~.sign) +
            scale_y_discrete(labels=function(x) str_wrap(x, width=60)) +
        ggtitle(paste0("dotplot for gene set enrichment of KEGG - ", save_name)))
  
   print(dotplot(reactome_obj, showCategory=20, split = ".sign", font.size = 9) + facet_grid(~.sign) +
            scale_y_discrete(labels=function(x) str_wrap(x, width=60)) +
        ggtitle(paste0("dotplot for gene set enrichment of REACTOME - ", save_name)))
          
}

```

# run enrichment and plotting


## within KI, high vs. low fat diet

KI_HFD-KI_Con.diet



```{r comp1, fig.height= 8, fig.width=10, fig.align = 'center' }

contrasts_list <- list("KI_HFD-KI_Con.diet",
                       "KI_HFD-WT_HFD",
                       "KI_Con.diet-WT_Con.diet",
                       "WT_HFD-WT_Con.diet")

contrasts_list <- set_names(contrasts_list, nm = contrasts_list)

read_DE_res_run_enrichment(res_df = read_tsv(paste0("res/edgeR_", contrasts_list[[1]], "_DE_results.tsv")), 
                           save_name = contrasts_list[[1]])


```

## within HFD, KI vs. WT

KI_HFD-WT_HFD

```{r comp2, fig.height= 8, fig.width=10, fig.align = 'center'}

read_DE_res_run_enrichment(res_df = read_tsv(paste0("res/edgeR_", contrasts_list[[2]], "_DE_results.tsv")), 
                           save_name = contrasts_list[[2]])
                           
                           
```

## within Con.diet, KI vs. WT

KI_Con.diet-WT_Con.diet

```{r comp3, fig.height= 8, fig.width=10, fig.align = 'center'}

read_DE_res_run_enrichment(res_df = read_tsv(paste0("res/edgeR_", contrasts_list[[3]], "_DE_results.tsv")), 
                           save_name = contrasts_list[[3]])
```

## within WT, high vs. low fat diet.

WT_HFD-WT_Con.diet
                   
```{r comp4, fig.height= 8, fig.width=10, fig.align = 'center'}

read_DE_res_run_enrichment(res_df = read_tsv(paste0("res/edgeR_", contrasts_list[[4]], "_DE_results.tsv")), 
                           save_name = contrasts_list[[4]])
```

